BOLD Parameter Estimation Using Particle Filters

### 2010-06-26 doc/thesis/thesis.tex
got q-q plots in

\documentclass{article}
\usepackage[left=3cm,top=2cm,right=3cm,nohead,nofoot]{geometry}

\usepackage{times}
\usepackage[siunitx]{circuitikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}
\usepackage{hyperref}

@@

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity for the majority of the brain.
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite its prevalence, there have been relatively few developments
in the actual analysis of FMRI images. A steady stream of studies have built
on the original BOLD signal derivation first described in \cite{Ogawa}, 
from the Baloon model first proposed by \cite{Buxton1998}
all the way to full fully autonomous system of equations \cite{Riera2004}. And while
there have been numerous forks in the model, enough in fact to make an entire paper
studying the differences,\cite{Deneux2006}, it is widely known that all these
models have quantitatively less bias error than General Linear Model which is
typically employed today. Then again, depending on the model there may be between
seven \cite{Riera2004} and 50 \cite{Behzadi2005} parameters per voxel to
be optimized. Clearly there is a significant risk of error due to variance
with so many degrees of freedom, not to mention a significantly increased
computation cost. In this thesis I demonstrate the use of a 
particle filter as a means of addressing these problems.

FMRI images as a method of detecting neural activation is based on 
temporal changes of the Blood Oxygen Level Dependent (BOLD) signal.
The BOLD signal is caused by minute changes in the ratio of Deoxygenated
Hemoglobin to Oxygenated Hemoglobin in blood vessels throughout the brain.
Because Deoxygenated hemoglobin is paramagnetic, higher concentrations
attenuate the signal when using T2 weighted imaging techniques, such 
as Echo Planar Imaging (EPI) which is used in FMRI. When axons becomes active,
a large amount of ions quickly flow out of the cell. In order for the action
potential to be used again, an active pumping process moves ions back into the
axon. This process of recharging the axon takes a large amout of enengy, which 
naturally uses oxygen. On a massive scale (cubic millimiter) this activation/recharge
process is happening all
the time; however, it happens at a much higher rate when a portion of
the brain is very active. Thus, blood vessels in a very active area will 
tend to have less oxygenated hemoglobin, and more deoxygenated hemoglobin,
resulting in lower FMRI signal. However, to compensate for activation, muscles that
control blood vessels relax in that region to allow more blow flow, which 
in fact results in a higher concentration of oxygenated hemoglobin. Thus,
increased activation actually tends to \emph{increase} the MR signal in
comparison with the base level. It is this overcompensation that is the 
primary signal detected with FMRI imaging. This cascade of events
can, as a consequence of increased activity, increase the local metabolism, 
blood flow, blood volume, and oxygenated hemoglobin; though not necessarily
in sync. The lag between these
various factors is what causes many of the complexities of the BOLD signal.

\subsection{FMRI}
Magnetic Resonance Imaging, MRI is a method of building 3D images
non-invasively, based on the difference between nuclear spin
relaxation times in various molecules. Initially the entity being
imaged is brought into a large magnetic field which aligns the spins
of molecules in the same direction; radio frequency (RF) signals may
then be used to excite nuclear spin away from the steady alignment. 
As the nuclei precess back to their original orientation, they resonate
at the same RF frequency of their original excitation. Conveniently, the
excitation of nuclear spins return their original state at different
rates, called the T1 relaxation time, depending on the properties of 
the material excited. Additionally, the
coherence of the spins also decay differently (and quite a bit faster
than T1) based on the properties of the region that has been excited.
This gives two primary methods of contrasting substances,
which is the basis of T1 and T2 weighted images. Additionally, there
dephasing occurs at two different rates, the T2 relaxation time,
which is impossible to recover from, and T2$^*$ relaxation, which is
much faster, but possible to recover from with an inversion pulse.
Oftentimes T1 relaxation times can be on the order of seconds if 
a significant excitation pulse is applied. 
In order to rapidly acquire entire brain images, as is done in Functional 
MRI, a single large excitation pulse is applied to the entire brain,
and the entire volume is acquired in a single T1 relaxation period. 
Because the entire k-space (spatial-frequency) volume is acquired 
from a single excitation, the signal to noise ration is very low
in this type  of imaging (Echo Planar Imaging).

Increasing the spatial resolution of EPI imaging necessarily 
requires more time or faster magnetic field switching. Increasing
magnet switching rates though is difficult, because it can result in
more artifacts, or even lower signal to noise ratios. The result is
that at \emph{best} FMRI is capable of 1 second temporal resolution. 
Additionally, the means that each voxel of the image will contain 
the sum of a large amount neurons, capillaries and veins. Thus, the
FMRI signal, which is sensitive to the chemical composition of 
materials, is summing up the composition of various types of tissue
in addition to the blood, whose composition is what we actually care about.
In particular, the presence of Deoxyhemoglobin, Hemoglobin whose
oxygen has been used by a metabolic process, has a decreased magnetic
response compared to Oxygenated Hemoglobin. Thus, capillaries near
very active cells will typically have a higher Deoxyhemoglobin content and
lower signal, and regions with lower activity will have a lower 
Deoxyhamoglobin content and thus higher signal.
Unfortunately, as mentioned previously, blood is only a small part of
each voxel, which means that a single EPI image doesn't tell much about
Deoxyhemoglobin content. However assuming blood is the only thing changing
in the short term, percent difference from a baseline signal \emph{will}
tell us something. FMRI analysis is thus necessarily performed on the percent change
from the baseline. Luckily the assumption that tissue content does not change in
the short run is actually pretty good, although other factors can pollute
the baseline signal, as we will discuss later.

\subsection{BOLD Physiology}
It is well known that the two types of hemoglobin act as a contrast agents in 
EPI imaging
\cite{Buxton1998}, \cite{WEISSKOFF1994}, \cite{Ogawa}, however the connection
between Deoxyhemoglobin/Oxygenated Hemoglobin and neural activity is non-trivial. 
Intuitively, increased 
metabolism will increase Deoxyhemoglobin, however blood vessels are quick
to compensate by increasing local blood flow. Increased inflow will of course
preceed increased outflow, and increased inflow is accomplished by loosening 
capilary beds. Both of these factors drive increased storage capacity.
Since the local MR signal depends on the ratio of Deoxyhemoglobin to Oxygenated
Hemoglobin, increased volume of blood can certainly effect this ratio if 
metabolism doesn't exactly match the increased inflow of oxygenated blood.
This was the impetus
for the ground breaking balloon model (\cite{Buxton1998}) and windkessel
model (\cite{Mandeville1999}). These models derive from first principals
the increased deoxyhemoglobin ratio and volume of capillaries based on a given flow.
These were the first two attempts to quantitatively account for the shape of the 
BOLD signal as a consequence of the lag between the cerebral blood volume (CBV) 
and the cerebral blood flow (CBF). In fact \cite{Buxton1998} went to far as
to show that a simple, well chosen blood flow waveform coupled with a square 
wave cerebral metabolic rate of oxygen (CMRO2) curve, in the context of a balloon 
model, could fully account for the BOLD signal.

Although \cite{Buxton1998} showed that a well chosen flow waveform could 
explain much of the BOLD signal, there was still a matter of proposing a
realistic waveform for the CBF and for the CMRO2. \cite{friston2000} gave
a reasonable and simple
expressoin for CBF input,$f$, based on a flow inducing signal, $s$, 
\begin{eqnarray}
\dot{s} &=& \epsilon u(t) - \frac{s}{\tau_s} - \frac{f - 1}{\tau_f} \\
\dot{f} &=& s
\end{eqnarray}
where $\epsilon$ is a neuronal efficiency term, $u(t)$ is a stimulus, and $\tau_f$, $\tau_s$
are both time constants. In \cite{Buxton2004} the final piece of the simple balloon
model was put into place, by describing the CMRO2 as a constant multiple of
the CBF (the inflow of blood). This completed the basic balloon model, and 
was well summarized in \cite{Riera2004}. 
\begin{eqnarray}
\dot{v} &=& \frac{1}{\tau_0}(f - v^\alpha)\\
\dot{q} &=& \frac{1}{\tau_0}(\frac{f(1-(1-E_0)^f)}{E_0} - \frac{q}{v^{1-1/\alpha}})
\end{eqnarray}
where $v$ is normalized cerebral blood volume (CBV), and $q$ is the normalized
local deoxyhemoglobin/oxygenated hemoglobin ratio, $E_0$ is the resting metabolic
rate and $\alpha$ is Grubb's parameter controling the balloon model. 
\cite{Obata2004} refined the readout equation of the BOLD signal based on the
deoxyhemoglobin content (q) and local blood volume (v), resulting in the
final BOLD equation:
\begin{eqnarray}
y   &=& V_0((k_1 + k_2)(1-q) - (k_2 + k_3)(1-v))\\
k_1 &=& 4.3 \times \nu_0 \times E_0 \times TE = 2.8\\
K_2 &=& \epsilon_0 \times r_0 \times E_0 \times TE = .57\\
k_3 &=& \epsilon_0 - 1 = .43
\end{eqnarray}
Where $\nu_0 = 40.3 s^{-1}$  is the frequency offset in Hz for fully
deoxygenated blood (at 1.5T), $r_0 = 25 s^{-1}$  is the slope relating
change in relaxation rate with change in blood oxygenation, and
$\epsilon_0 = 1.43$ is the 
ratio of signal MR from intravascular to extravascular at rest. Although,
obviously these constants change with experiment ($TE$, $\nu_0$, $r_0$),
patient, and brain 
region ($E_0$, $r_0$), often the estimated values taken from \cite{Obata2004} are used
as constants ($k_1 + k_2 = 3.4$, and $k_2+k_3 = 1$) in 1.5 Tesla studies..
While this model is in a sense complete, it is far from perfect. The major
problem often brought up with this version of the BOLD model is that it
does not represent the so called "post-stimulus undershoot" well.
The post-stimulus undershoot is the name for a prolonged sub-normal
BOLD response for a period of 10 to 60 seconds after stimulus has
ceased (\cite{Chen2009}, \cite{Mandeville1999a}).

There are two theories for the cause of the post stimulus undershoot. Recall
that a lower than base signal means that there is an increased deoxyhemoglobin
content in the voxel. The first and simplest explanation is that the post-stimulus
undershoot is caused by a prolonged increase in CMRO2 after CBV and CBF
have returned to their base levels. This theory is justified by quite a few
studies that show CBV and CBF returning to the baseline before the BOLD signal
(\cite{Frahm2008}, \cite{Donahue2009}, \cite{Buxton2004}, \cite{Lu2004},
\cite{Shen2008}). Unfortunately, because of limitations on FMRI and en vivo
CBV/CBF measurement techniques it is difficult to isolate whether CBF and
CBV truly have returned to their baseline. Other research seems to indicate
that there can be a prolonged residual supernormal CBV (\cite{Mandeville1999a}, 
\cite{Behzadi2005}, \cite{Chen2009a}), although none of these papers completely
rule out the possiblity of increased CMRO2. Additionally, in \cite{Yacoub2006}, it 
was found that the post-stimulus undershoot varried across regions of the brain, 
which could further explain the contradictions found elsewhere. \cite{Chen2009a}
makes a compelling case that most of the post stimulus undershoot could be 
explained be a prolonged CBV increase, and a prolonged CBF undershoot, and that
many of the previous measurements showing a quick recovery of CBV 
may have been dominated arterial CBV's return to baseline.

Because of the significant possility of a completely independent CMRO2,
extremely complex models for metabolism exist (\cite{Zheng2005}), although 
most recent studies have
focused on their ability to explain the prolonged BOLD post stimulus 
undershoot \cite{Zheng2005}, \cite{Buxton2004}. This is because \cite{Buxton2004}
and later \cite{Riera2004} showed that the main portion of the signal may be 
accurately estimated by a simple blood flow locked expression of the CMRO2.. 
Although \cite{Deneux2006} did not deal extensively with prolonged post
stimulus undershoot, the comparisons made in that publication showed minimal
improvement from separate expressions of CMRO2, in comparison to the much 
increased complexity. \cite{Deneux2006} did show that by simply adding viscoelastic
terms, first proposed in \cite{Buxton2004}, that a slowed return to baseline for
the BOLD signal is possible to model. However, viscoelastic effects primarily
control CBV, which, as mentioned already, many studies have claimed cannot be 
responsible for the BOLD post-stimulus undershoot. Another extensive model
that attempts to quantify the post-stimulus undershoot is the compliance model 
proposed by \cite{Behzadi2005}. Although through a somewhat different means than
the \cite{Zheng2005} and \cite{Buxton2004} papers, its possible the increased
model flexibility ultimately is the key reason for the improvements, as opposed
to increased plausibility. Because of these controversies, and because this is
the first time a particle filter has been used for this problem in this way, 
our aim to keep the model simple was best met by using the original balloon
model with the possible addition of the visco-elastic effects from \cite{Buxton2004}.

Even more advanced versions of the Balloon model exist. In fact \cite{Buxton2004}
introduced several additional state variables, including the CMRO2, the O2 extraction
fraction, which is closely related to CMR02, and the neural response, which 
causes the stimulus to decay toward some steady state value. The 
neural response is intended to emulate neural habituation, wherein neurons
become less sensitive to a prolonged stimulus. While these advanced may be more
capable of capturing a more exact version of the BOLD signal, the difference
between the model will often times be below the noise floor. In essence this 
is a classic bias-variance
dilemma: at some point increased model flexibility, and thus variance, is not
worth the decrease in model bias. For now it remains to be seen where this
line may be drawn in the BOLD signal, although \cite{deneux2006} does not
show a significant improvement from the additional parameters added in 
\cite{Buxton2004}.

\subsection{Previous Studies of Parameters}
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. In that paper, Friston et. al. used a variation of
Expectation Maximization to find a normal distribution for the parameters:
\begin{eqnarray}
\epsilon &=& N(.54 , .1 ^2 )  \nonumber \\
\tau_s & =&  N(1.54, .25^2)   \nonumber \\
\tau_f & =&  N(2.46, .25^2)   \nonumber \\
\tau_0 & =&  N(.98 , .25^2 )   \nonumber \\
\alpha & =&  N(.33 , .45^2 )   \nonumber \\
E_0   & =&  N(.34 ,  .1 ^2 )   \nonumber \\
V_0  & = &  .03 (not\ estimated) \nonumber
\end{eqnarray}
Since then, several other methods of have been used to calculate
BOLD parameters from FMRI timecourses. In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is in some since the
"right" way to find the solution, it comes with several caveats. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations could go on
indefinitely. Second, it still assumes the parameters and noise are Gaussian, and
will only be optimal in that case. Third, depending on the nonlinearities present
in the system, local minima may be extremely common. Later \cite{Hu2009} used an 
Unscented Kalman Filter over all the parameters and state variables to find the 
parameter set/variable time series. While this method has the drawback of not necessarily
being optimal, unfortunately there is no general optimal solution to non-linear non
Gaussian models. Hu et. al.'s technique also will run significantly faster than
ML based techniques, since it does not require recalculating the entire timeseries
for every step in parameters. Both \cite{Hu2009} and \cite{Friston2002} came to results 
very similar to the expected values stated in \cite{Buxton1998} and \cite{Friston2000}.
One potential problem with all these techniques are that the depend heavily on the
priors. The starting point of the parameters could have a huge impact on the
results, and while one can be hopeful that this isn't the reason for the agreement
between \cite{Friston2000} and later results, there is no way to know.

In \cite{Jonston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static (classicly called
parameters) and dynamic parameters (classically known as state variables).
Essentially a particle filter is used to calculate the state variables at some
time, and then the estimated distribution of the particles was used to find
the most likely set of parameters that would give that distribution of state variables.
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} guesses:
\begin{eqnarray}
epsilon &=& .069 \pm .014    \nonumber \\
tau_s & =& 4.98 \pm 1.07  \nonumber \\
tau_f & =& 8.31 \pm 1.51   \nonumber \\
tau_0 & =& 8.38 \pm 1.5    \nonumber \\
alpha & =& .189 \pm .004   \nonumber \\
E_0   & =& .635 \pm .072     \nonumber \\
V_0   & =& .0149 \pm .006     \nonumber 
\end{eqnarray}
Notably, the time constants are significantly longer. This could be a result of
some preprocessing to the stimulus timeseries performed in \cite{Friston2002} and
later works but not in \cite{Johnston2008}, or it could be that \cite{Johnston2008}
depended less on the priors.

Its possible, although unlikely that the balloon model is not possible to learn
without detrimental cost in variance error. I do not think this is the case, however,
and I will work to dispell this possibility in the results with simulated time
series.

\subsection{Noise}
Thus, despite some discrepencies, the cause of the BOLD signal is
relatively well known.  but FMRI doesn't detect this happening in one neuron,
but rather as the 
aggregate over millions of cells. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation or noise. 
A particularly insidious type of noise present in FMRI is a low frequncey
drift, first characterized by a Weiner process in \cite{Riera2004}. 
Though not present in all regions, it is prevalent enough to cause significant
inference problems \cite{Tanabe2002}. It is still not
clear where exactly this noise comes from, although it is possible it is 
the result of magnets heating up, or some distortion in magnetic
fields \cite{Smith2007}. It is clear that this drift signal is not soley
due to a physiological
effects, given its presence in cadavers and phantoms \cite{Smith1999}.
Additionally, since it is standard operating procedure to perform
coregistration between volumes, it is unlikely that movement was
the cause of the drift either (otherwise the effect would not be 
widely reported).

\section{Current Techniques}
\subsection{Basic Statistical Parametric Mapping}
Although not strictly the same thing as parameter calculation from 
FMRI, activation detection is very similar. In fact, estimation of 
parameters is somewhat a generalization of the idea of activation detection.
Thus, it is important to draw a comparison between the methods proposed
in this thesis with existing methods of activation detection.

The most basic method of analyzing FMRI data is through a standard T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
period during which the stimulus was activate separately then treating 
them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the probability that the samples came from the same distribution
(the null hypothesis). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
50000 or more samples is on average going to generate $.05*50000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection{General Linear Model}
The most used form of FMRI analysis is Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus (see \cite{Hofmann1997}). By using hierarchical models
the output signal timeseries is considered the weighted
sum of the various input timeseries. Essentially every experimental factor
is considered as another level inputs (ex. which patient is an input). 
The equation for a general linear model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of input, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the inputs plus some error. The calculation
of $\beta$ then performed using a maximum likelihood or gradient descent search 
to minimize the error.

<Image of GLM>

It is well known of course that a square wave stimulus does not result in a square wave
in the activation of brain regions. Thus, various methods are used to 
smooth $X(t)$'s related to actual input (as opposed to experimental confounds
such as patient number, age etc) through time, and bandlimit the input. 
The best technique is convolving the stimulus input with a hemodynamic 
response function, which mimicks the basic shape of BOLD activation, including a delay
due to rise time and fall time. The downside of this method, however is that 
the hemodynamic signal is static, meaning the same hemodynamic function is
used for every region of the brain, and $Y(t)$ must be a constant
scalar multiple of $X(t)$ to be appropiately identified.

<Image of Hemodynamic Response Function>

The GLM is extremely powerful at determining the linear dependence of
a set of regressors on the output. Unfortunately, there is significant evidence
that many of these dependencies are nonlinear, which means they may be
difficult or impossible to detect using linear techniques. Additionally, 
the lack of flexibility in delay variables certainly hinders SPM's
ability to locate voxels that do not conform to the canonical hemodynamic 
response function. The very fact that different hemodynamic response functions
work better for different regions/experiments is a testament to the need
for increased flexibility.

Linear models also ignore prior knowledge about the system. Combined FMRI
CBF or CBV imaging methods are available and could shed much light on
neural activation; however there is no way of combining that data into a
linear model. Although the most advanced form of the general linear model includes
a Hemodynamic Response Function, it is static across every region of the brain;
yet it is well known that capillary beds are not uniform and so blood perfusion 
cannot possibly be static across the brain. Its possible that activation is
every bit as strong in regions that aren't found by SPM, but that there are
nonlinearities that do not match the canonical HRF; resulting in signficant bias error. 
It is not uncommon
for data to be thrown out in FMRI studies because no significant activation
has been seen. This practice highlights the fact that strictly linear
approaches are insufficiently flexible to account for relatively common
variations of neural activation.

<Image with two different $\alpha$s>

<image comparing the results of 10\% changes in various signals>

In general activation detection type methods also don't have the ability 
to find pathologies based on state variables or parameters.  It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now.

Finally, all these techniques require the noise to be Gaussian to reach
an optimal solution. In fact there is no known optimal solution to
nonlinear models with non-Guassian noise, so obviously some assumptions
are going to have to be made to reach a solution. However, it would be
nice to have an algorithm that is robust to these effects, and could
still give a good solution when Gaussianity is violated.

\section{Unscented Kalman Filter}
The unscented kalman filter is a powerful gaussian/bayes filter that attempts
to model the posterior distribution of dynamical systems as a multivariate
Gaussian. The Unscented Kalman Filter (UKF) generalizes the Extended Kalman
Filter by allowing the state update to be a function, $g$,

\begin{eqnarray}
X(t) = g(u(t), X(t-1))\\
Y(t) = h(X(t)
\end{eqnarray}

In order to estimate the posterior at $t$, a deterministic set of "sigma" points 
(often 2 per dimension, plus 1 at the mode of the multivariate distribution)
weighted according to a Gaussian estimate of $X(t-1)$ are passed through
the update equation. This set of points are then used to estimate the 
mean and covariance of $X(t)$. The benefit of this, is that it requires
no Jacobian and only a few extra calculations to get a decent esimate of
a posterior Gaussian. In the BOLD case, the set of equations we are modeling have no 
closed form solution, and finding the Jacobian is impossible without approximations. Although 
\cite{Riera2004} claims to use a Jacobian for their solution, they are
in fact using the Jacobian of $\dot{s}$, $\dot{q}$, $\dot{f}$, $\dot{v}$,
rather than $s$, $q$, $f$, and $v$. This is important because the Extended
Kalman filter depends on the jacobian to map a gaussian through the 
advancement of time. Thus the popular Extended Kalman Filter won't work
in this case, whereas the Unscented Kalman Filter still does.

\begin{figure}
\label{fig:EKFWorking}
\caption{Examples of update distributions, using an Kalman Filter, \cite{Thrun}}
\end{figure}

The difficulty
of using a Kalman Filter, however, is that it assumes a multivariate 
Gaussian for the state variables, $X(t-1)$. The problem with this is that 
when a system is significantly nonlinear, the state at 
$X(t)$ will almost certainly be non-Gaussian, and thus estimating
$X(t+1)$ with $X(t)$ as a multivariate Gaussian will perpetually introduce
error in the distribution. Furthermore, it is not really known what 
sort of underlying distributions may exist in such a mixed biological,
mechanical, chemical system such as the brain. Assuming the parameters
all to be Gaussian may in fact be a gross error. On the other hand, if
the our tests indicate a Gaussian distribution is a good fit, it would
be a wise plan to go back and consider the use of a modified Unscented 
Kalman Filter rather
than the full Particle Filter we discuss in the next section. These are 
non-trivial issues given that the assumption of Gaussianity is what allows
the UKF to estimate the posterior using only the first and second moments;
two parameters that don't uniquely describe most distributions.

Another potential problem with the UKF is that typically the sigma points,
used to estimate the posterior probability, $P(X(t) | X(t-1), u(t))$, are
located on the main axes. As a result, the covariances are not allowed to 
be effected by the state transition the same way the variances are. While
this may be reasonable in low-dimensional systems, high demsional systems
have a much greater potential for interplay between the variables. While
this problem is relatively easy to fix (by using more sigma points off
the main axes), there is a definite cost in complexity.

Ultimately, the nonlinearities of the BOLD model make Gaussian estimates
too unrealistic for practical use. As \ref{fig:Transformation} shows, even 
holding all other parameters steady, the nonlinearities due to $q$ and $v$
alone are enough to give a non-gaussian result from a two variables 
gaussian input.

\begin{figure}

\label{fig:Transformation}
\caption{a) A Two-Variables gaussian input of $q$ and $v$, b) The 
resulting distribution after being evolved for two seconds. c) The
estimated Gaussian based on the mean and covariance of b). d) The
residual between the two distributions.}
\end{figure}

\section{Proposed Approach}
\subsection{Goal}
The ultimate goal of this project is to provide a new set of tools
for analyzing FMRI data. Whereas SPM techniques have been highly 
successful at finding macroscopic regions of activation, linear 
modeling can carry significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect is mitigated by the fact
that we plan to use a model that is based on first principals. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that one or more
parameters are a suitable replacement for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

\subsection{Introduction to Particle Filters}
Particle filters, a type of Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters give a timeseries of measurements. Unlike Markov 
Chain Monte Carlo estimation, Sequential Monte-Carlo methods are designed
to be used with parameters that vary with time. Unlike variations of the
Kalman filter, particle filters do not make the assumption that noise
is Gaussian. Thus particle filters are often the best solution to bayesian 
tracking for non-linear, non-gaussian systems. 
%todo: might need to reference attempts to model a system of our type with
%unscented kalman filter
\subsubsection{Model}
The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF can introduce bias; however, this effect can be
minimized by alterring the initial weights in the mixture pdf. Of course
every gradient descent must choose starting points and it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose a set or stream of measurements
are given, $\{y(t), t = 1, 2, 3, ... T\}$, where $T$ is permitted to go
 to infinity. Then the goal is to find the 
parameters, $\hat{\theta}$, and underlying state time series, $\hat{x}[0:T]$
that minimize the difference between $\hat{y}[0:T]$
and $y[0:T]$. In our case, we will assume that we know the form
of the model, which is based on first principals, and that
there is some true $\theta$ and a true time-series of underlying
state variable, $x[0:T]$ that drives $y[0:T]$. Assuming a model form 
such as we do here reduces model variance, potentially at the cost of increased
bias (or systematic) error. We will assume a basic state space model:

\begin{equation}
\dot{x}(t) = f(t, x(t), u(t), \theta, \nu_x)
\end{equation}

\begin{equation}
y(t) = g(t, x(t), u(t), \theta, \nu_y)
\end{equation}

Where $x(t)$ is a vector of state variables, $\theta$ is a vector of system
constants, $u(t)$ is a stimulus, $y(t)$ an observation, and
$\nu_x$ and $\nu_y$ are random variates. Obviously any one of these could
be a vector, so for instance $u(t)$ could encode multiple types of stimuli.

Although not strictly necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, the systems are assumed to be time invariant. This 
assumption is based on the idea that if you froze the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happend. 
Few biological systems are predictable enough for them to be summarized
by a time varying function. Even systems that one might assume work in that
way are actually much more complicated. The heart may seem like an obvious
exception, however, the period between heartbeats vary often enough that prediction
would necessitate another state-space model. In short, we
assume no parameters are time varying, because not enough information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, so unanticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Third, we will assume noise is additive, and that $\nu_x$ may be projected into
a Weiner, or other summing process that is additive with $g$ and $\nu_y$, which
will be named $\nu_d$.
Finally, $x(t)$ will encapsulate $\theta$, the unknown model constants, which
means that the vector $\dot{x}$ will always have members
that are 0. The results of these assumptions are a simplified version of the
state space equations:

\begin{equation}
\label{stateass}
\dot{x}(t) = f(x(t), u(t))
\end{equation}

\begin{equation}
\label{measass}
y(t) = g(x(t)) + \nu_y + \nu_d
\end{equation}

Because $\nu_d$ is something akin to and additive Weiner process $y[0:T]$, it 
will include low frequency noise. $\nu_y$ on the other hand will cause i.i.d. noise
in $y[0:T]$. For some of the tests, I will use de-trending methods to reduce the effects of 
$\nu_d$, the remainder of which will be re-factored into $\nu_y$. Both $\nu_d$ and $\nu_y$
have biological and non-biological sources. MR can lead to both types of noise, 
as demonstrated in \cite{drift}. Meanwhile changes in metabolism, heart rate, or
other biochemical intervention could all lead to either $\nu_d$ or $\nu_y$.

\subsubsection{Prior}

The goal of the particle filter is to evolve a probably distribution 
$Pr(\hat{x}(T) | u[0:T], y[0:T])$,
that asymptotically approaches the probability distribution $Pr(x(T) | u[0:T])$.
Considering that $y$ contains measurement noise as well as noise from $x$,
 it is clear that $Pr(x(t) | u[0:T])$ is not a single true value
but a true posterior. 
To begin with, the particle filter starts with a prior distribution, and $N_p$
particles need to be drawn from that distribution, $\alpha(x)$:
\begin{equation}
\{\hat{Pr}x_i(0),w_i] : x_i(0) \sim \alpha(x), w_i = \frac{1}{N_p}, i \in \{1, 2, ... , N_p\} \}
\end{equation}

Where $N_p$ is the number of particles or points used to describe the prior 
using a Mixture PDF. 
\begin{equation}
\hat{Pr}(x(0) = \hat{x}) = \sum_{i=1}^{N_p} w_i\delta(\hat{x} - x_i(0) ) dx
\end{equation}
Where $\delta(x-x_0)$ is 1 if and only if $x = x_0$ (the Kronecker delta function).

If a true prior is preferred, then the weights
should all be $1/N_p$, and since $x_i$ was drawn from the prior, this will
be an approximation of the prior distribution. If a relatively flat prior is 
preferred, then each particle's weight could be divided by the density, $\alpha(x_i)$,
which creates a flat prior with support points in the region of $\alpha(x)$. Either
way, $\alpha(x)$ should be much broader than the true posterior, $Pr(x(0))$, since the
choice of support points is crucial to the convergence of any sampling importance
algorithm. 
For the BOLD signal all the parameters have been studied and have relatively well
known mean and variance, so a prior could be very helpful. We ran simulations for
both normalized and un-normalized priors, although we believe in cases such as this,
where a good prior exists, it should be used. 
For strictly positive parameters (members of $x$) we used a gamma distribution,
whereas for parameters that could be negative, we used a Gaussian distribution. In
both cases standard deviations twice that found in previous studies were used.

Note that all the probabilities implicitly depend on $u[0:T]$, so those terms 
will be left off for simplicity.
Once the probability, $\hat{Pr}(x(T) | x[0:T-1], y[0:T-1])$ has been found
(initially this is just Mixture approximating the prior since no measurements are 
available and no previous probabilities are available), its possible to approximate
the probability for short times between times when measurement is available, by shifting
the probability according the progression of the state equations. This is only 
an approximate, since integrating $\nu_d$ should increase uncertainty as
time without a measurement passes.

\begin{equation}
\hat{Pr}(x(T+\Delta t)) \approx 
\sum_{i=1}^{N_p} w_i\delta\left(x - (x_i(T) + \int_T^{T+\Delta} \dot{x}_i(t) dt) \right)
\end{equation}

\subsubsection{Weighting}
When a measurement becomes available it is incorporated into the probability.
This process of incorporating new data is called sequential importance sampling,
and eventually causes the probability to converge. The weight is defined
as

\begin{equation}
\label{weightfunc}
w_i(T) \propto \frac{\hat{Pr}(x_i[0:T] | y[0:T])}{q(x_i[0:T] | y[0:T])}
\end{equation}

>>>>

\documentclass{report}
\usepackage[left=3cm,top=2cm,right=3cm]{geometry}

\usepackage{times}
\usepackage[siunitx]{circuitikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{cancel}
\usepackage[colorlinks=true]{hyperref}

@@

\begin{document}
\title{BOLD Parameter Estimation Using Particle Filters}
\author{Micah Chambers\\Virginia Polytechnic Institute and State University}
\date{}
\maketitle

\tableofcontents

\include{intro}
\include{background}
\include{particlefilters}
\include{methods}
\include{results}
\include{conclusion}
